---
title: "Viral Load Time Series Segmentation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Viral Load Time Series Segmentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `PvlTimeSeriesSegmentation` package provides tools for analyzing viral load time series data by identifying basins (stable states) and bridges (transitions) using kernel density estimation in 2D feature space.

## Basic Usage

```{r setup, eval=FALSE}
library(PvlTimeSeriesSegmentation)
```

### Generate Sample Data

```{r sample-data, eval=FALSE}
# Generate sample viral load data with decline-rebound pattern
sample_data <- generate_sample_data(12, pattern = "decline_rebound")
head(sample_data)
```

### Interpolate to Feature Space

```{r interpolate, eval=FALSE}
# Convert to (value, slope) feature space
interpolated <- interpolate_viral_load(sample_data, n_points = 30)
head(interpolated)
```

### Segment Time Series

```{r segment, eval=FALSE}
# Identify basins and bridges using KDE
segmentation <- segment_time_series(interpolated)

# View summary
segmentation$summary
```

### Classify States

```{r classify, eval=FALSE}
# Add descriptive state labels
classified <- classify_states(segmentation)

# View state distribution
table(classified$data$state_label)
```

### Visualize Results

```{r plot, eval=FALSE}
# Create trajectory plot in (value, slope) space
plot_segmentation(classified, plot_type = "trajectory")

# Create time series plot with segmentation
plot_segmentation(classified, plot_type = "time_series")

# Create density contour plot
plot_segmentation(classified, plot_type = "density")
```

## Algorithm Details

1. **Feature Extraction**: Time series data is interpolated and slopes are calculated
2. **Kernel Density Estimation**: 2D KDE identifies density patterns in (value, slope) space
3. **Basin/Bridge Classification**: High-density regions become basins, low-density paths become bridges
4. **Variance Heuristics**: Additional variance-based criteria refine bridge identification
5. **Temporal Constraints**: Minimum basin sizes ensure temporal stability

## Parameters

- `density_threshold`: Separates basins (above) from bridges (below)
- `min_basin_size`: Minimum consecutive points for a valid basin
- `variance_threshold`: Variance-based criterion for bridge identification

## Use Cases

- HIV viral load monitoring during treatment
- Disease progression analysis
- Treatment efficacy assessment
- Viral rebound detection